A Meta-analysis of Longevity Estimates of Mosquito Vectors of Disease: Supplementary Material

### 2018-11-23 supplementary_materials.tex
Some references added and read through to `Number of gonotrophic cycles estimates from dissection' subsection
Read up until discussion
Going through supplementary up Collection of dissection data section

\section{MRR experiments}
\subsection{Data}
The MRR database in \cite{guerra2014global} contains 393 individual time-series, along with meta-data for a range of factors for each experiment (for example, species, geography, and date of study). We  cleaned the data, making a number of amendments to it (see Section \ref{sec:amendments}), and transformed the data into a form amenable to our statistical model. We removed time-series with fewer than six separate recapture observations, resulting in 238 time series. For the species-level analyses, we required at least two time series per species, resulting in loss of six series, meaning that the data used in the Bayesian hierarchical analysis comprised 232 individual time series (see Section \ref{sec:appendix_mrrStudyList} for a list of the studies included). Tables \ref{tab:mrr_aggregateData} and \ref{tab:mrr_IndividualData} summarise the data used in the Bayesian hierarchical analysis. This data encompassed time-series from MRR experiments across 33 different species and three genera: \textit{Aedes} (91 separate time-series), \textit{Anopheles} (94), and \textit{Culex} (47), and spanning a wide geographical range (Figure \ref{fig:mrr_lifetimes_map}). The data comprise 177 female-only, 35 male-only and 18 mixed-sex release time-series. Where the age of the adult mosquitoes at release-time was available (102 time-series) we use this to increment our time variable (that represents mosquito age) in the statistical models. Where age was unavailable (130 time-series), we assume that the age at release is zero. We recognise this assumption will likely underestimate mosquito age for some studies and so we interpret our estimates as lower bounds on mosquito lifespan.

@@

\begin{table}[htbp]
	\centering
	\begin{tabular}{lc}
		\toprule
		\textbf{Data characteristic} & \textbf{Total} \\
		\midrule
		Number of species & 33 \\
		Number of genera & 3 \\
		Number of \textit{Anopheles} & 94 \\
		Number of \textit{Aedes} & 91 \\
		Number of \textit{Culex} & 47 \\
		Number of female-only MRR & 179 \\
		Number of male-only MRR & 35 \\
		Number of mixed sex MRR & 18 \\
		Number of pre-release blood-feeding only MRR & 71 \\
		Number of pre-release sugar-feeding only MRR & 41 \\
		Number of pre-release both blood- and sugar-feeding MRR & 4 \\
		Number of pre-release neither blood- and sugar-feeding MRR & 116 \\
		Number of MRR time-series & 232 \\
		\bottomrule
	\end{tabular}%
	\caption{Summary of variables across all MRR time-series.}
	\label{tab:mrr_aggregateData}%
\end{table}%

\begin{table}[htbp]
	\centering
	\footnotesize
	\begin{adjustwidth}{-0.5in}{-0.5in}%
		\begin{tabularx}{1.25\textwidth}{l|ccccc}
			\toprule
			\textbf{Data characteristic} & \textbf{Min} & \textbf{Mean} & \textbf{Median} & \textbf{Max} & \textbf{Standard deviation} \\
			\midrule
			Study duration, days & 6.0   & 11.8  & 10.0  & 71.0    & 4.7 \\
			Number of days on which collections took place & 6.0   & 10.1  & 9.0   & 47    & 9.1 \\
			Number of separate release days & 1.0     & 1.9   & 1.0     & 23    & 3.0 \\
			Number released & 66    & 4,929  & 1,297  & 86,200 & 12,043 \\
			Number recaptured & 2     & 163   & 63    & 4,090  & 399 \\
			Recapture percentage & 0\% & 8.6\% & 5.2\% & 57.1\% & 10.04\% \\
			Age at release, days & 0   & 1.8   & 0   & 13    & 2.8 \\
			\bottomrule
		\end{tabularx}%
		\caption{Summary of data from individual MRR time-series.}
		\label{tab:mrr_IndividualData}%
	\end{adjustwidth}
\end{table}%

@@

\subsection{Statistical model}\label{sec:mrr_statistical}
Data for a typical MRR experiment consists of a single release of $N_{R}$ marked mosquitoes followed by a series of marked mosquito recaptures on subsequent days (Figure \ref{fig:mrr_exampleMRRSeries} shows three such example series.) We model the number $y(t)$ of marked mosquitoes recaptured on day $t$ using a negative binomial sampling model,
%
\begin{equation}\label{eq:NB}
y(t) \sim \text{NB}\left((N_{R} - Y(t-1)) S(t) \psi, \kappa\right),
\end{equation}
%
where $Y(t-1)$ is the cumulative number of mosquitoes caught on all days before $t$, $S(t)$ is the probability that an individual mosquito survives and remains in the study area until time $t$, $\psi$ is the daily recapture probability for an individual mosquito which is assumed constant through time, and $\kappa$ is the time-independent shape parameter of the negative binomial distribution that controls the extent of variance in recapture rate likely due mostly to environmental heterogeneity. We chose a parameterisation of the negative binomial such that its mean is given by $\mu(t) = (N_{R} - Y(t-1)) S(t) \psi$, and its variance by $\sigma(t)^2 = \mu(t) + \frac{\mu(t)^2}{\kappa}$.

@@

The simplest assumption is that death and dispersal rates are independent of mosquito age, giving an exponential `survival' function,
%
\begin{equation}
S(t) = e^{-\lambda t},
\end{equation}
%
where $\lambda$ is the sum of the rates of death and dispersal from the study area. To estimate the lifespan of mosquitoes we use this model because we find no evidence in support of age-dependence across all species (see below for the method used to determine this, and later the `Results').

@@

\subsection{Individual time-series estimates}\label{sec:MRR_individual_analysis}
We first treat each time-series separately, and estimate individual $(\lambda,\psi,\kappa)$ parameters of the statistical model (described in the previous section) for each time-series. To use a Bayesian methodology we must specify priors on this set of parameters. Here we choose to specify independent priors on each parameter of the form: $\lambda\sim \mathcal{N}(-2.32,1)$, $\psi\sim \text{exp}(50)$ (with $\psi$ constrained to be less than 0.1) and $\kappa\sim \textnormal{log-normal}(2,1)$. This prior on the rate parameter of the exponential distribution ($\lambda$) corresponds to a wide range of possible lifespans (Figure \ref{fig:mrr_individualTimeSeries_priors}A), with a mean of 10 days. It was necessary to use an informative prior on $\psi$ to allow it to be estimated; the prior on $\kappa$ is fairly uninformative and allows a wide range of values (Figure \ref{fig:mrr_individualTimeSeries_priors}B,C).

@@

\subsection{Estimating lifespan at the species, genus and overall groupings}\label{sec:MRR_hierarchical}
We synthesise information from across all MRR experiments to produce more robust estimates of mosquito lifespan than can be obtained from considering the individual time-series separately. However there exists considerable heterogeneity across the experiments. This heterogeneity has two sources. There is that arising from variability in experimental methodology. However there is also variability from actual differences in lifespan across the different mosquito cohorts; for example due to genetic differences between mosquito populations or due to climatic differences. Because of this heterogeneity we use a Bayesian hierarchical model which is akin to a `random effects' model in classical statistics. This type of model assumes that there is random variation in parameters at the individual time-series level, although each of the parameters is drawn from a common `population-level' distribution. In our case we separately estimate three different model groupings where the population-level distributions correspond to the species, genus and overall (across all studies) levels respectively (Figure \ref{fig:mrr_methodDiagramHierarchy}). This allows us to estimate average mosquito lifespan for a given species or genus, or alternatively across all time-series, by independent sampling from the posterior predictive distribution from which the individual $\lambda_i$ are drawn.

@@

Whilst we assume that the variation in mosquito mortality across experiments is random in nature, we recognise that there may exist systematic sources of variation that we do not include in our model. To account for this potential source of bias we examined how the individual estimates of lifespan correlated with experimental covariates (number of traps, range of traps, trapping method), but found no such evidence of systematic variation in lifespan (data not shown).

A Bayesian framework requires that we incorporate our pre-analysis beliefs into our estimates through the use of priors for all parameters in a particular model. In our hierarchical model we are required to specify prior distributions at two levels of the analysis. The first of these links the individual time-series estimates with the overarching group-level distributions. The parameters of these group-level distributions are then set prior distributions (Figure \ref{fig:mrr_methodDiagramHierarchy}). As an example, we suppose that the rate parameter $\lambda_i$ of our exponential survival model in experiment $i$ is given by:
%
\begin{align}
\lambda_i = \text{exp}(&c_i + \delta^m_{\text{genus}[i]} \text{sex}^{\text{m}}_i + \delta^{\text{mix}}_{\text{genus}[i]} \text{sex}^{\text{mix}}_i +\\  & \delta^{\text{sugar}}_{\text{genus}[i]} \text{sugar}_i +  \delta^{\text{blood}}_{\text{genus}[i]} \text{blood}_i),
\end{align}
%
where $\delta^m_{\text{genus}[i]}$ measures a genus-specific effect of male-only releases ($\text{sex}^{\text{m}}_i=1$) versus female-only ($\text{sex}^{\text{m}}_i=0$); $\delta^{\text{mix}}_{\text{genus}[i]}$ is the analogous effect for mixed releases versus female-only; $\delta^{\text{sugar}}_{\text{genus}[i]}$ is the analogous effect for mosquitoes that were sugar-fed prior to release versus those that were not fed; and $\delta^{\text{blood}}_{\text{genus}[i]}$ is the analogous effect for mosquitoes that were blood-fed prior to release versus those that were not fed. The parameter $c_i$ is modelled hierarchically, being drawn from a group-level distribution assumed to be log-normal,
%
\begin{equation}
c_i \sim  \textnormal{normal}(\mu_\lambda,\sigma)
\end{equation}
%
We then specify priors on these group-level parameters, $\mu_\lambda \sim N(-2.32,1)$ and $\sigma \sim \textnormal{log-normal}(-3,1)$. On all $\delta$ parameters, except $\delta^{\text{mix}}_{\text{genus}[i]}$, we specify standard normal priors. We specify a uniform prior on $\delta^{\text{mix}}_{\text{genus}[i]}\sim U(0,\delta^m_{\text{genus}[i]})$ so that the effect of mixed releases lies between the female-only release effect (0) and the male-only release effect.

The relatively complex nature of priors in hierarchical models make it important to determine their influence on inferences. We chose the above -- somewhat uninformative -- priors to allow a range of mosquito lifespans in order to minimise their effect on the estimates we report (Figure \ref{fig:mrr_meanLifespanPrior}). We also chose to set hierarchical priors on the remaining parameters in our models -- $\psi$ the probability of daily recapture, and $\kappa$ the `over-dispersion' parameter of a negative binomial distribution. For $\psi$ we chose an informative prior that placed all probability mass $\psi\leq 10\%$, since in most experiments the fraction of mosquitoes captured was significantly below this value. For $\kappa$ we chose a fairly wide prior that had most of its probability mass below $\kappa=20$ (Figure \ref{fig:mrr_PsiKappaPriors}). In all cases the hierarchical priors were chosen to have similar implications on lifespan, recapture probability and over-dispersion at the individual time-series level as for the non-hierarchical analysis described previously.

@@

The parameters of the survival function of each model were assigned hierarchical priors that allowed considerable variation in mosquito lifespan (see Figure \ref{fig:mrr_meanLifespanPrior}). Otherwise the statistical model we used remained the same as for the constant mortality case. In testing for age-dependent mortality, we did not account for sex or pre-release feeding effects due to the complexity of including these elements in more complex survival models.

@@

\begin{table}
	\footnotesize
	\noindent\makebox[\textwidth]{%
		\begin{tabularx}{1.2\textwidth}{l|c|X}
			\textbf{Survival function} & \textbf{time-series-level priors} & \textbf{Group-level priors}\\
			\midrule
			Exponential & $\lambda\sim \textnormal{log-normal}(\mu_\lambda,\sigma)$ &$\mu_\lambda \sim N(-2.32,1)\text{,}\; \sigma\sim \textnormal{log-normal}(-3,1)$\\
			\midrule
			Gompertz & $\alpha,\beta\sim$ log-normal$(\mu_{\alpha|\beta},0.2)$ &$\mu_\alpha\sim N(-3,1)\text{,}\; \mu_\beta\sim N(-3,1.1)$\\
			\midrule
			Weibull & $\alpha,(\beta-1)\sim$  log-normal$(\mu_{\alpha|\beta},0.2)$ &$\mu_\alpha\sim N(-4.8,1.75)\text{,}\; \mu_\beta\sim N(-4,2)$\\
			\midrule
			Gompertz-Makeham & $\alpha,\beta,c\sim$ log-normal$(\mu_{\alpha|\beta|c},0.2)$ &$\mu_\alpha\sim N(-3,1)\text{,}\; \mu_\beta\sim N(-3,0.5)\text{,}\; \mu_c\sim N(-4.5,0.5)$\\
			\midrule
			Logistic & $\alpha,\beta,s\sim$  log-normal$(\mu_{\alpha|\beta|s},0.2)$ &$\mu_\alpha\sim N(-2.8,1)\text{,}\; \mu_\beta\sim N(-3,1)\text{,}\; \mu_s\sim N(-3,1)$\\
			\midrule
			Logistic-Makeham & $\alpha,\beta,s,c\sim$ log-normal$(\mu_{\alpha|\beta|s|c},0.2)$ &$\mu_\alpha\sim N(-3.2,1)\text{,}\; \mu_\beta\sim N(-3.2,1)\text{,}\; \mu_s\sim N(-3,0.5)\text{,}\; \mu_c\sim N(-4,0.5)$\\
			\bottomrule
	\end{tabularx}}\caption{\textbf{The priors used on parameters of each different survival model.} For the exponential model the `group-level' priors were the same for the genus and `overall' models that were also estimated. The exponential model was the only model that was simple enough to allow the scale parameter of the log-normal ($\sigma$) to be estimated by the data. The notation $\alpha,\beta\sim \text{log-normal}(\mu_{\alpha|\beta},0.2)$ means that $\alpha$ and $\beta$ were assigned independent log-normal priors with location parameters $\mu_\alpha$ and $\mu_\beta$ respectively, and a scale parameter of 0.2 in both cases.}\label{tab:mrr_priors}
\end{table}

\subsection{Model estimation by MCMC}\label{sec:mrr_MCMC}
The likelihood and priors we use result in posterior distributions whose analytic form cannot be calculated with existent computational methods. Instead we use Markov Chain Monte Carlo (MCMC) methods to sample from each posterior distribution. In particular we used \textit{Stan} software \citep{carpenter2016stan} that implements an efficient MCMC algorithm known as NUTS \citep{hoffman2014no}.

@@

\subsection{K-Fold cross validation}\label{sec:mrr_kFold}
One way to estimate a model's out-of-sample predictive capability is to use Akaike Information Criterion, Bayesian Information Criterion, Deviance Information Criterion or Watanabe Akaike Information Criterion that explicitly penalise a model in accordance to its complexity (typically indexed by its number of free parameters) to correct for model over-fit. The way in which this correction takes places however, is fairly heuristic, and less appropriate for hierarchical models. An alternative method, common in the machine learning literature, is known as `cross-validation' (see for example, \cite{kohavi1995study}). Here to estimate out of sample predictive capability, the original data set is partitioned into training and test sets. The model is then fitted to the training set, and its predictive performance measured on the independent test set.

@@

For each individual series we recorded the following meta-data: genus, species, collection method, whether or not insecticide was mentioned as being used in or around the time of mosquito collections, and the start and end dates of the experiment. At the article level we recorded the following meta-data: author, year, title, country, location (within country), start and end date, whether insecticide was used during any of the experimental replicates, whether a mrr experiment occurred alongside the dissection study, and the collection location (indoor or outdoor). At either the individual series or article levels we record additional meta-data describing the nature of data collection, for example explaining where the data was contained within the article, whether it was obtained by digitising graphs, and the number of separate dated series. For those few cases ($n=16$ series) where the data was obtained by digitising graphs, we used the WebPlotDigitizer online tool \citep{digitise}.

The data collection method resulted in 568 separate dissection series, across 72 published articles (see Section \ref{sec:appendix_dissectionStudyList} for a list of the studies included in our final dataset). The published datasets cover the period from 1960-2015 with comparable numbers of studies across each decade (Figure \ref{fig:dissection_timeSeries}). The statistical approach applied to the data relies on the assumption that there is a constant rate of recruitment into the adult population (see Section \ref{sec:dissection_dissectionStats}). If populations fluctuate strongly from month-to-month this assumption will likely be violated. To try to mitigate against such a risk, we aggregate the data across all dates to obtain a single series for each identifier, resulting in 201 such series. To obtain a reasonable level of accuracy on estimates we remove all those individual (aggregated) series where there are fewer than 100 mosquitoes in total in the series. Finally we remove data for any species where there was only one series resulting in 131 series across four genera (\textit{Anopheles}, \textit{Culex}, \textit{Aedes}, and \textit{Mansonia}) and 25 species (Table \ref{tab:dissection_speciesNumbers}). These studies are distributed across a wide range of geographies (Figure \ref{fig:dissection_map}). The raw data for the dissection analysis are shown in Figures \ref{fig:dissection_exampleData1}, \ref{fig:dissection_exampleData2} and \ref{fig:dissection_exampleData3}.

@@

Here we estimate our model with one of six different survival functions (the same as for the MRR case; Table \ref{tab:mrr_survivalDescription}), each of which makes different assumptions regarding how the force of mortality is affected by mosquito age. To estimate lifespan at the species, genus and overall groupings we then use a hierarchical Bayesian model of the same mathematical form as for the analysis of MRR experiments (see Section \ref{sec:MRR_hierarchical}). However the priors for the analysis of dissection data were modified so that they represented a mean prior lifespan of three gonotrophic cycles, although allowed considerable variation in mean lifespan (Figure \ref{fig:dissection_lifespanPriorsParity}). The priors used for the over-dispersion parameter ($\kappa$) for the negative binomial likelihood were the same as for the MRR analysis.

@@

Since we collected data on the method used to estimate the duration of the gonotrophic cycle we show the raw estimates for the two most common approaches: MRR and laboratory (Figure \ref{fig:dissection_gonotrophicCycleRaw_MRRVsLab}). From these results it is evident that the estimates from the laboratory studies are, on average, higher than those using the MRR method, (although we wait until Section \ref{sec:dissection_gonotrophicDuration} for a quantitative comparison).

>>>>

\section{MRR experiments}
\subsection{Data}
The MRR database in \cite{guerra2014global} contains 393 individual time-series, along with meta-data for a range of factors for each experiment (for example, species, geography, and date of study). We  cleaned the data, making a number of amendments to it (see Section \ref{sec:amendments}), and transformed the data into a form amenable for our statistical model. We removed time-series with fewer than six separate recapture observations, resulting in 238 time series. For the species-level analyses, we required at least two time series per species, resulting in loss of six series, meaning that the data used in the Bayesian hierarchical analysis comprised 232 individual time series (see Section \ref{sec:appendix_mrrStudyList} for a list of the studies included). Tables \ref{tab:mrr_aggregateData} and \ref{tab:mrr_IndividualData} summarise the data used in the Bayesian hierarchical analysis. This data encompassed time-series from MRR experiments across 33 different species and three genera: \textit{Aedes} (91 separate time-series), \textit{Anopheles} (94), and \textit{Culex} (47), and spanning a wide geographical range (Fig. \ref{fig:mrr_lifetimes_map}).

@@

\begin{table}[htbp]
	\centering
	\begin{tabular}{lc}
		\toprule
		\textbf{Data characteristic} & \textbf{Count} \\
		\midrule
		Species & 33 \\
		Genera & 3 \\
		\textit{Anopheles} & 94 \\
		\textit{Aedes} & 91 \\
		\textit{Culex} & 47 \\
		Female-only MRR & 179 \\
		Male-only MRR & 35 \\
		Mixed sex MRR & 18 \\
		Pre-release blood-feeding only MRR & 71 \\
		Pre-release sugar-feeding only MRR & 41 \\
		Pre-release both blood- and sugar-feeding MRR & 4 \\
		Pre-release neither blood- and sugar-feeding MRR & 116 \\
		MRR time-series & 232 \\
		\bottomrule
	\end{tabular}%
	\caption{\textbf{Summary of variables across all MRR time-series.} For the `Species' and `Genera' rows, the counts represent the number of distinct entries in the database; otherwise the counts represent the number of time series.}
	\label{tab:mrr_aggregateData}%
\end{table}%

\begin{table}[htbp]
	\centering
	\footnotesize
	\begin{adjustwidth}{-0.5in}{-0.5in}%
		\begin{tabularx}{1.25\textwidth}{l|ccccc}
			\toprule
			\textbf{Data characteristic} & \textbf{Min} & \textbf{Mean} & \textbf{Median} & \textbf{Max} & \textbf{Standard deviation} \\
			\midrule
			Study duration, days & 6.0   & 11.8  & 10.0  & 71.0    & 4.7 \\
			Number of days on which collections took place & 6.0   & 10.1  & 9.0   & 47    & 9.1 \\
			Number of separate release days & 1.0     & 1.9   & 1.0     & 23    & 3.0 \\
			Number released & 66    & 4,929  & 1,297  & 86,200 & 12,043 \\
			Number recaptured & 2     & 163   & 63    & 4,090  & 399 \\
			Recapture percentage & 0\% & 8.6\% & 5.2\% & 57.1\% & 10.04\% \\
			Age at release, days & 0   & 1.8   & 0   & 13    & 2.8 \\
			\bottomrule
		\end{tabularx}%
		\caption{\textbf{Summary of data from individual MRR time-series.}}
		\label{tab:mrr_IndividualData}%
	\end{adjustwidth}
\end{table}%

@@

\subsection{Statistical model}\label{sec:mrr_statistical}
Data for a typical MRR experiment consists of a single release of $N_{R}$ marked mosquitoes followed by a series of marked mosquito recaptures on subsequent days (Fig. \ref{fig:mrr_exampleMRRSeries} shows three such example series). We model the number $y(t)$ of marked mosquitoes recaptured on day $t$ using a negative binomial sampling model,
%
\begin{equation}\label{eq:NB}
y(t) \sim \text{NB}\left((N_{R} - Y(t-1)) S(t) \psi, \kappa\right),
\end{equation}
%
where $Y(t-1)$ is the cumulative number of mosquitoes caught on all days before $t$, $S(t)$ is the probability that an individual mosquito survives and remains in the study area until time $t$, $\psi$ is the daily recapture probability for an individual mosquito which is assumed constant through time, and $\kappa$ is the time-independent shape parameter of the negative binomial distribution that controls the extent of variance in recapture rate likely due mostly to environmental heterogeneity. We chose a parameterisation of the negative binomial such that its mean is given by $\mu(t) = (N_{R} - Y(t-1)) S(t) \psi$, and its variance by $\sigma(t)^2 = \mu(t) + \frac{\mu(t)^2}{\kappa}$.

@@

The simplest assumption is that death and dispersal rates are independent of mosquito age, giving an exponential `survival' function,
%
\begin{equation}
S(t) = e^{-\lambda t},
\end{equation}
%
where $\lambda$ is the sum of the rates of death and dispersal from the study area. To estimate the lifespan of mosquitoes we use this model, partly because of the limited evidence we determine in favour of senescence (Figs. \ref{fig:mrr_elpd} and \ref{fig:dissection_elpd}).

@@

\subsection{Individual time-series estimates}\label{sec:MRR_individual_analysis}
We first treat each time-series separately, and estimate individual $(\lambda,\psi,\kappa)$ parameters of the statistical model (described in the previous section) for each time-series. To use a Bayesian methodology we must specify priors on this set of parameters. Here we choose to specify independent priors on each parameter of the form: $\lambda\sim \mathcal{N}(-2.32,1)$, $\psi\sim \text{exp}(50)$ (with $\psi$ constrained to be less than 0.1, a threshold below which the majority of daily recapture probabilities in the data are observed) and $\kappa\sim \textnormal{log-normal}(2,1)$. This prior on the rate parameter of the exponential distribution ($\lambda$) corresponds to a wide range of possible lifespans (Fig. \ref{fig:mrr_individualTimeSeries_priors}A), with a mean of 10 days. It was necessary to use an informative prior on $\psi$ to bound it to sensible values and to prevent the correlation of estimates with $\lambda$ leading to unreasonably short estimates of $\lambda$; the prior on $\kappa$ is fairly uninformative and allows a wide range of values (Fig. \ref{fig:mrr_individualTimeSeries_priors}B,C).

@@

\subsection{Estimating lifespan at the species, genus and overall groupings}\label{sec:MRR_hierarchical}
We synthesise information from across all MRR experiments to produce more robust estimates of mosquito lifespan than can be obtained from considering the individual time-series separately. However there exists considerable heterogeneity across the experiments. This heterogeneity has two sources. There is that arising from variability in experimental methodology. There is also variability from actual differences in lifespan across the different mosquito cohorts; for example due to genetic differences between mosquito populations or due to climatic differences. Because of this heterogeneity we use a Bayesian hierarchical model which is akin to a `random effects' model in classical statistics. This type of model assumes that there is random variation in parameters at the individual time-series level, although each of the parameters is drawn from a common `population-level' distribution. In our case we separately estimate three different model groupings where the population-level distributions correspond to the species, genus and overall (across all studies) levels respectively (Fig. \ref{fig:mrr_methodDiagramHierarchy}). This allows us to estimate average mosquito lifespan for a given species or genus, or alternatively across all time-series, by independent sampling from the posterior predictive distribution from which the individual $\lambda_i$ are drawn.

@@

Whilst we assume that the variation in mosquito mortality across experiments is random in nature, we recognise that there may exist systematic sources of variation that we do not include in our model. To account for this potential source of bias we examined how the individual estimates of lifespan correlated with experimental covariates (trapping method and number or density of traps). The data on trapping method in the Guerra et al., (2014) database is less reliable than other data entries because published MRR studies may use multiple trapping methods and typically do not disaggregate captures by method. When we examined the correlation of trapping method with lifespan estimates, however, we did not find systematic variation by method (data not shown). We determined a positive correlation between trapping density and lifespan estimates (Fig. \ref{fig:mrr_lifeSpanVsTrapDensity}), which indicates the impact of mosquito dispersal on estimated lifespan.

A Bayesian framework requires that we incorporate our pre-analysis beliefs into our estimates through the use of priors for all parameters in a particular model. In our hierarchical model we are required to specify prior distributions at two levels of the analysis. The first of these links the individual time-series estimates with the overarching group-level distributions. The parameters of these group-level distributions are then set prior distributions (Fig. \ref{fig:mrr_methodDiagramHierarchy}). As an example, we suppose that the rate parameter $\lambda_i$ of our exponential survival model in experiment $i$ is given by:
%
\begin{align}
\lambda_i = \text{exp}(&c_i + \delta^m_{\text{genus}[i]} \text{sex}^{\text{m}}_i + \delta^{\text{mix}}_{\text{genus}[i]} \text{sex}^{\text{mix}}_i +\\  & \delta^{\text{sugar}}_{\text{genus}[i]} \text{sugar}_i +  \delta^{\text{blood}}_{\text{genus}[i]} \text{blood}_i),
\end{align}
%
where $\delta^m_{\text{genus}[i]}$ measures a genus-specific effect of male-only releases ($\text{sex}^{\text{m}}_i=1$) versus female-only ($\text{sex}^{\text{m}}_i=0$); $\delta^{\text{mix}}_{\text{genus}[i]}$ is the analogous effect for mixed releases versus female-only; $\delta^{\text{sugar}}_{\text{genus}[i]}$ is the analogous effect for mosquitoes that were sugar-fed prior to release versus those that were not fed; and $\delta^{\text{blood}}_{\text{genus}[i]}$ is the analogous effect for mosquitoes that were blood-fed prior to release versus those that were not fed. The parameter $c_i$ is modelled hierarchically, being drawn from a group-level distribution assumed to be normal,
%
\begin{equation}
c_i \sim  \mathcal{N}(\mu_\lambda,\sigma)
\end{equation}
%
We then specify priors on these group-level parameters, $\mu_\lambda \sim \mathcal{N}(-2.32,1)$ and $\sigma \sim \textnormal{log-normal}(-3,1)$. On all $\delta$ parameters, except $\delta^{\text{mix}}_{\text{genus}[i]}$, we specify standard normal priors. We specify a uniform prior on $\delta^{\text{mix}}_{\text{genus}[i]}\sim U(0,\delta^m_{\text{genus}[i]})$ so that the effect of mixed releases lies between the female-only release effect (0) and the male-only release effect ($\delta^m_{\text{genus}[i]}$).

The relatively complex nature of priors in hierarchical models make it important to determine their influence on inferences. We chose the above -- somewhat uninformative -- priors to allow a range of mosquito lifespans in order to minimise their effect on the estimates we report (Fig. \ref{fig:mrr_meanLifespanPrior}). We also chose to set hierarchical priors on the remaining parameters in our models -- $\psi$ the probability of daily recapture, and $\kappa$ the `over-dispersion' parameter of a negative binomial distribution. For $\psi$ we chose an informative prior that placed all probability mass $\psi\leq 10\%$, since in most experiments the fraction of mosquitoes captured was significantly below this value. For $\kappa$ we chose a fairly wide prior that had most of its probability mass below $\kappa=20$ (Fig. \ref{fig:mrr_PsiKappaPriors}). In all cases the hierarchical priors were chosen to have similar implications on lifespan, recapture probability and over-dispersion at the individual time-series level as for the non-hierarchical analysis described previously.

@@

The parameters of the survival function of each model were assigned hierarchical priors that allowed considerable variation in mosquito lifespan (see Fig. \ref{fig:mrr_meanLifespanPrior}). Otherwise the statistical model we used remained the same as for the constant mortality case. In testing for age-dependent mortality, we did not account for sex or pre-release feeding effects due to the complexity of including these elements in more complex survival models.

@@

\begin{table}
	\footnotesize
	\noindent\makebox[\textwidth]{%
		\begin{tabularx}{1.2\textwidth}{l|c|X}
			\textbf{Survival function} & \textbf{time-series-level priors} & \textbf{Group-level priors}\\
			\midrule
			Exponential & $c\sim \mathcal{N}(\mu_\lambda,\sigma)$ &$\mu_\lambda \sim N(-2.32,1)\text{,}\; \sigma\sim \textnormal{log-normal}(-3,1)$\\
			\midrule
			Gompertz & $\alpha,\beta\sim$ log-normal$(\mu_{\alpha|\beta},0.2)$ &$\mu_\alpha\sim N(-3,1)\text{,}\; \mu_\beta\sim N(-3,1.1)$\\
			\midrule
			Weibull & $\alpha,(\beta-1)\sim$  log-normal$(\mu_{\alpha|\beta},0.2)$ &$\mu_\alpha\sim N(-4.8,1.75)\text{,}\; \mu_\beta\sim N(-4,2)$\\
			\midrule
			Gompertz-Makeham & $\alpha,\beta,c\sim$ log-normal$(\mu_{\alpha|\beta|c},0.2)$ &$\mu_\alpha\sim N(-3,1)\text{,}\; \mu_\beta\sim N(-3,0.5)\text{,}\; \mu_c\sim N(-4.5,0.5)$\\
			\midrule
			Logistic & $\alpha,\beta,s\sim$  log-normal$(\mu_{\alpha|\beta|s},0.2)$ &$\mu_\alpha\sim N(-2.8,1)\text{,}\; \mu_\beta\sim N(-3,1)\text{,}\; \mu_s\sim N(-3,1)$\\
			\midrule
			Logistic-Makeham & $\alpha,\beta,s,c\sim$ log-normal$(\mu_{\alpha|\beta|s|c},0.2)$ &$\mu_\alpha\sim N(-3.2,1)\text{,}\; \mu_\beta\sim N(-3.2,1)\text{,}\; \mu_s\sim N(-3,0.5)\text{,}\; \mu_c\sim N(-4,0.5)$\\
			\bottomrule
	\end{tabularx}}\caption{\textbf{The priors used on parameters of each different survival model.} For the exponential model the `group-level' priors were the same for the genus and `overall' models that were also estimated. The exponential model was the only model that was simple enough to allow the scale parameter of the log-normal ($\sigma$) to be estimated by the data. The notation $\alpha,\beta\sim \text{log-normal}(\mu_{\alpha|\beta},0.2)$ means that $\alpha$ and $\beta$ were assigned independent log-normal priors with location parameters $\mu_\alpha$ and $\mu_\beta$ respectively, and a scale parameter of 0.2 in both cases.}\label{tab:mrr_priors}
\end{table}

\subsection{Model estimation by MCMC}\label{sec:mrr_MCMC}
The likelihood and priors we use result in posterior distributions whose analytic form cannot be calculated with existent computational methods. Instead we use Markov Chain Monte Carlo (MCMC) methods to sample from each posterior distribution. We used \textit{Stan} software \citep{carpenter2016stan} that implements an efficient MCMC algorithm known as NUTS \citep{hoffman2014no}.

@@

\subsection{K-Fold cross validation}\label{sec:mrr_kFold}
One way to estimate a model's out-of-sample predictive capability is to use the Akaike Information Criterion that explicitly penalises a model in accordance to its complexity (typically indexed by its number of free parameters) to correct for model over-fit. The way in which this correction takes places however, is fairly heuristic, and less appropriate for hierarchical models. An alternative method, common in the machine learning literature, is known as `cross-validation' (see for example, \cite{kohavi1995study}). Here to estimate out of sample predictive capability, the original data set is partitioned into training and test sets. The model is then fitted to the training set, and its predictive performance measured on the independent test set.

@@

For each individual series we recorded the following meta-data: genus, species, collection method, whether or not insecticide was mentioned as being used during or shortly before the time of mosquito collections, and the start and end dates of the experiment. At the article level we recorded the following meta-data: author, year, title, country, location (within country), start and end date, whether insecticide was used during any of the experimental replicates, whether a mrr experiment occurred alongside the dissection study, and the collection location (indoor or outdoor). At either the individual series or article levels we record additional meta-data describing the nature of data collection, for example explaining where the data was contained within the article, whether it was obtained by digitising graphs, and the number of separate dated series. For those few cases ($n=16$ series) where the data was obtained by digitising graphs, we used the WebPlotDigitizer online tool \citep{digitise}.

The data collection method resulted in 568 separate dissection series, across 72 published articles (see Section \ref{sec:appendix_dissectionStudyList} for a list of the studies included in our final dataset). The published datasets cover the period from 1960-2015 with comparable numbers of studies across each decade (Fig. \ref{fig:dissection_timeSeries}). The statistical approach applied to the data relies on the assumption that there is a constant rate of recruitment into the adult population (see Section \ref{sec:dissection_dissectionStats}). If populations fluctuate strongly from month-to-month this assumption will likely be violated. To try to mitigate against such a risk, we aggregate the data across all dates to obtain a single series for each identifier, resulting in 201 such series. To obtain a reasonable level of accuracy on estimates we remove all those individual (aggregated) series where there are fewer than 100 mosquitoes in total in the series. Finally we remove data for any species where there was only one series resulting in 131 series across four genera (\textit{Anopheles}, \textit{Culex}, \textit{Aedes}, and \textit{Mansonia}) and 25 species (Table \ref{tab:dissection_speciesNumbers}). These studies are distributed across a wide range of geographies (Fig. \ref{fig:dissection_map}). The raw data for the dissection analysis are shown in Figures \ref{fig:dissection_exampleData1}, \ref{fig:dissection_exampleData2} and \ref{fig:dissection_exampleData3}.

@@

Here we estimate our model with one of six different survival functions (the same as for the MRR case; Table \ref{tab:mrr_survivalDescription}), each of which makes different assumptions regarding how the force of mortality is affected by mosquito age. To estimate lifespan at the species, genus and overall groupings we then use a hierarchical Bayesian model of the same mathematical form as for the analysis of MRR experiments (see Section \ref{sec:MRR_hierarchical}). However the priors for the analysis of dissection data were modified so that they represented a mean prior lifespan of three gonotrophic cycles, although allowed considerable variation in mean lifespan (Fig. \ref{fig:dissection_lifespanPriorsParity}). The priors used for the over-dispersion parameter ($\kappa$) for the negative binomial likelihood were the same as for the MRR analysis.

@@

Since we collected data on the method used to estimate the duration of the gonotrophic cycle we show the raw estimates for the two most common approaches: MRR and laboratory (Fig. \ref{fig:dissection_gonotrophicCycleRaw_MRRVsLab}). From these results it is evident that the estimates from the laboratory studies are, on average, higher than those using the MRR method, (although we wait until Section \ref{sec:dissection_gonotrophicDuration} for a quantitative comparison).
